---
title: "HW11: Education Level and Survey Weighting"
author: "Divya Vakkalanka"
---

```{r setup, include=FALSE}
library(kableExtra)
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
```

# Introduction

In this assignment, you'll examine how survey weighting affects our understanding of voting intentions across different education levels, using the data we examined in class. You'll calculate both weighted and unweighted statistics, create visualizations, and reflect on the implications for reporting.

## The Data

The nonvoters dataset contains survey responses about voting intentions and behaviors from a national survey. The survey was conducted prior to an election and includes demographic information like education level.

```{r}
# Load the dataset
nonvoters_data <- read_csv("https://raw.githubusercontent.com/dwillis/jour405_files/refs/heads/main/nonvoters_data.csv")

# Take a quick look at the data structure
glimpse(nonvoters_data)
```

### Key Variables

- `weight`: Survey weight assigned to each respondent
- `Q21`: Voting intention (1 = Yes, 2 = No, 3 = Unsure/Undecided)
- `educ`: Education level (College, Some college, High school or less)

## Task 1: Education Distribution

First, let's examine the distribution of education levels in our sample. Replace "REPLACE_ME" with the correct variable for education level.

```{r}

education_distribution <- nonvoters_data |>
  count(educ) |>
  mutate(percentage = n / sum(n) * 100) |>
  kable(digits = 1, col.names = c("Education Level", "Count", "Percentage (%)"))

education_distribution
```

## Task 2: Reflection Question

Why might education levels in survey samples often differ from the general population? What factors might cause certain education groups to be over or underrepresented?

people with higher education levels might be more likely to fill out surveys, overrepresenting them in the results. 


## Task 3: Unweighted Analysis by Education

Now, let's calculate unweighted voting intentions by education level. This is what we would report if we didn't apply any weighting to our sample.

```{r}
# Calculate unweighted voting intentions by education
unweighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Count responses
  summarize(count = n(), .groups = "drop_last") |>
  # Calculate percentages
  mutate(total = sum(count),
         percentage = count / total * 100) |>
  ungroup()

# Create a more readable format with voting intentions as columns
unweighted_educ_summary <- unweighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(unweighted_educ_summary, digits = 1, caption = "Unweighted Voting Intentions by Education Level")
```

## Task 4: Reflection Question

Based just on this unweighted analysis, what headline might you write for a news story about education and voting intentions?

College Students More Likely to Vote 
## Task 5: Weighted Analysis by Education

Next, let's apply survey weights to see how this changes our results. Instead of just counting responses, we'll sum the weights for each group. Replace "REPLACE_ME" with the appropriate weight variable

```{r weighted-by-education}

weighted_by_education <- nonvoters_data |>
  # Filter out missing values
  filter(!is.na(Q21), Q21 > 0, !is.na(educ)) |>
  # Group by education and response
  group_by(educ, Q21) |>
  # Sum the weights instead of counting
  summarize(weighted_count = sum(weight), .groups = "drop_last") |>
  # Calculate weighted percentages
  mutate(weighted_total = sum(weighted_count),
         weighted_percentage = weighted_count / weighted_total * 100) |>
  ungroup()

# Create a more readable format
weighted_educ_summary <- weighted_by_education |>
  pivot_wider(
    id_cols = educ,
    names_from = Q21,
    values_from = weighted_percentage,
    names_prefix = "pct_"
  ) |>
  rename(
    "Yes (%)" = pct_1,
    "No (%)" = pct_2,
    "Unsure (%)" = pct_3
  )

kable(weighted_educ_summary, digits = 1, caption = "Weighted Voting Intentions by Education Level")
```

## Task 6: Reflection Questions

1. How did the percentages change after applying weights? Which education group showed the biggest changes?

After weighting the results, it better represented all groups meaning that the groups that had education levels of high school or less were less likely to vote. 

2. Why might the weighted results be considered more accurate than the unweighted results?

The weighted results are more accurate than unweighted because they can be representative better for the whole population since there is less sampling biases.

## Task 7: Comparison of Weighted vs. Unweighted Results

Let's create a direct comparison table to see the differences more clearly.

```{r}

comparison <- unweighted_educ_summary |>
  inner_join(weighted_educ_summary, by = "educ", suffix = c("_unweighted", "_weighted")) |>
  mutate(
    # Calculate the differences between weighted and unweighted percentages
    yes_diff = `Yes (%)_weighted` - `Yes (%)_unweighted`,
    no_diff = `No (%)_weighted` - `No (%)_unweighted`,
    unsure_diff = `Unsure (%)_weighted` - `Unsure (%)_unweighted`
  ) |>
  # Select just the columns we want to display
  select(educ, yes_diff, no_diff, unsure_diff) |>
  rename(
    "Education Level" = educ,
    "Yes (% point diff)" = yes_diff,
    "No (% point diff)" = no_diff,
    "Unsure (% point diff)" = unsure_diff
  )

kable(comparison, digits = 1, caption = "Difference Between Weighted and Unweighted Results (percentage points)")
```

## Task 8: Reflection Question

Which education group shows the largest differences between weighted and unweighted results?

The group that shows the largest differences is the high school or less with a 2.7 point decrease in yes and 1.5 increase in no. 

## Task 9: Visualization

Visualizations can help us see the differences more clearly. Let's create a bar chart comparing weighted and unweighted "Yes" responses by education level. Replace "REPLACE_ME" with the correct variable name

```{r}
educ_viz_data <- bind_rows(
  # Unweighted data
  unweighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses (Q21=1)
    mutate(Type = "Unweighted") |>
    select(Type, educ, percentage),
  
  # Weighted data - 
  weighted_by_education |> 
    filter(Q21 == 1) |>  # Only "Yes" responses
    mutate(
      Type = "Weighted",
      percentage = weighted_percentage
    ) |>
    select(Type, educ, percentage)
)

# Create a grouped bar chart
ggplot(educ_viz_data, 
       aes(x = educ, y = percentage, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(aes(label = sprintf("%.1f%%", percentage)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(
    title = "Weighted vs. Unweighted 'Yes' Responses by Education",
    subtitle = "Q21: Do you plan to vote in the November election?",
    y = "Percentage (%)",
    x = "Education Level"
  ) +
  scale_fill_manual(values = c("Unweighted" = "#619CFF", "Weighted" = "#F8766D")) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    legend.position = "bottom"
  )
```

## Task 10: Reflection Questions

Does the visualization make it easier to see the differences between weighted and unweighted results? Why or why not?

The visualization helps us see the differences because it shows us the side by side differences where it is the most significant. 
## Task 11: Summary

Based on your analysis of weighted and unweighted results by education level, write a brief (2-3 paragraph) journalistic summary of what you found. You should try to address:

1. How education level relates to voting intentions
2. How weighting affected your understanding of this relationship
3. What this means for interpreting polling results in news reporting

Those with higher education levels have a higher intention to vote in the November election. 93.5% of those with a College Education are likely to vote in the election. This is compared to 70% who are likely to vote if they have a high school education or less and 85.1% of those with some college education. 

Weighting the data was important to understand how more college educate survey respondents were found meaning they could have been overrepresented in the data. Once the survet was weighted, this bias was not as significant and ensured that no group was overexaggerated. 

This is important in news reporting so that conclusions are not hastily drawn and show inaccurate differences between groups who answer the survey. 

## Task 12: Final Reflection Questions

1. Why is it important for journalists to understand survey weighting when reporting on polls?
Polls being inaccurately reported on is a concern if journalists do not understand survey weighting. 

2. How might the differences between weighted and unweighted results affect how you would report on this data?
Weighting might change how significant the gaps are between groups. 

3. What additional information would you want to know about how the weights were calculated before using this data in a news story?
What were the different weights used to calculate?