---
title: "HW7_MandMs"
---

In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse
```{r}
library(tidyverse)
```


### Task 2: Load the M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandm <- read_csv ("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10.
```{r}
sample_5 <- mandm |>
  sample_n(5)

sample_10 <- mandm |>
  sample_n(10)
```


### Task 4: Calculate the mean and standard deviation of the samples, using one of the colors.

```{r}
sample_10 |> summarize(mean = mean(red), sd = sd(red))
sample_5 |> summarize(mean = mean(red), sd = sd(red))
```

### Task 5: Calculate the mean and standard deviation of the population, using the same color as Task 4

```{r}
mandm |> summarize(mean = mean(red), sd = sd(red))
```

### Task 6: Compare the sample and population statistics

Compare the three sets of means and standard deviations. Are they similar?
The sample taken from 5 and 10 were very similar with the means only being .3 apart but a significant difference in the standard deviation with a difference of 10. For the population, the difference was a lot more with the population mean being 72.98 compared to the samples which had 75 and the standard deviation was 16.46 which is in between the standard deviations the samples produced. 

### Task 7: Describe the results

Write a few sentences about what you found in the data. What does this tell you about the M&Ms dataset and about the act of random sampling?

This tells us that given the sample set we get in this population, the data could vary since there are a decent amount of outliers so the standard deviation can be high depending on what is chosen in the random sample. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
